### 2023-01-24 latex/table1.tex
converted to peerj overleaf template

This manuscript
(\href{https://stain.github.io/2022-fdo-paper/v/86bc7fe899c7464fad8f4ebe872efb61c3be1394/}{permalink})
was automatically generated
from \href{https://github.com/stain/2022-fdo-paper/tree/86bc7fe899c7464fad8f4ebe872efb61c3be1394}{stain/2022-fdo-paper@86bc7fe}
on January 24, 2023.

\hypertarget{authors}{%
\section*{Authors}\label{authors}}
\addcontentsline{toc}{section}{Authors}

\begin{itemize}
\item
  \textbf{Stian Soiland-Reyes}
  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
  \href{https://orcid.org/0000-0001-9842-9718}{0000-0001-9842-9718}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
  \href{https://github.com/stain}{stain}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
  \href{https://twitter.com/soilandreyes}{soilandreyes}
  Department of Computer Science, The University of Manchester, UK; Informatics Institute, Faculty of Science, University of Amsterdam, NL
  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
\item
  \textbf{Carole Goble}
  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
  \href{https://orcid.org/0000-0003-1219-2137}{0000-0003-1219-2137}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
  \href{https://github.com/carolegoble}{carolegoble}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
  \href{https://twitter.com/CaroleAnneGoble}{CaroleAnneGoble}
  Department of Computer Science, The University of Manchester, UK
  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://www.eosc-life.eu/}{EOSC-Life} (European Commission \href{https://cordis.europa.eu/project/id/824087}{H2020-INFRAEOSC-2018-2 824087}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{https://biodt.org/}{BioDT} (European Commission \href{https://cordis.europa.eu/project/id/101057437}{HORIZON-INFRA-2021-TECH-01-01 101057437}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
\item
  \textbf{Paul Groth}
  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
  \href{https://orcid.org/0000-0003-0183-6910}{0000-0003-0183-6910}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
  \href{https://github.com/pgroth}{pgroth}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
  \href{https://twitter.com/pgroth}{pgroth}
  Informatics Institute, Faculty of Science, University of Amsterdam, NL
\end{itemize}

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

\textbf{TODO}: Rewrite from notes below to actual abstract.

FAIR Digital Object is an emerging concept from EOSC. This is important. Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape. Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.

What is the message of this paper?

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  These are the overlaps
\item
  This is what FDO is requiring but not commonly deployed
\item
  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
\item
  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
\item
  Lessons for Semantic Web community. What is missing in FDO and in SW.
\item
  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
\item
  Go back to the old SW layer cake. Explains why we picked these frameworks.
\item
  Lessons for SW: Parts of the stack that is less. FDO contributions.
\end{enumerate}

Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.

Have all the ingredients, but not cooking properly.

Systematic through frameworks.

Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?

\hypertarget{sec:introduction}{%
\section*{Introduction}\label{sec:introduction}}

The FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}) encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures. They have been widely recognised by the research community and policy makers as a goal to strive for. In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures (\protect\hyperlink{ref-u8Cy0psL}{Mons et al., 2017}). The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.

Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures. The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.

In other discourse, Linked Data (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\protect\hyperlink{ref-EJDjHrUv}{Bonino Da Silva Santos et al., 2016}; \protect\hyperlink{ref-GJHYN6xW}{Hasnain \& Rebholz-Schuhmann, 2018}). Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}), opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\protect\hyperlink{ref-99MS5xE0}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021a}; \protect\hyperlink{ref-J2H8yssV}{Neumann, Laranjeiro \& Bernardino, 2021}). While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022}).

One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\protect\hyperlink{ref-uxWjmlio}{Stefi, 2015}; \protect\hyperlink{ref-v0WwFHgT}{Stefi \& Hess, 2015}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves (\protect\hyperlink{ref-5V3GqSzR}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021b}).

These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data. Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.

Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.

\hypertarget{sec:background}{%
\section*{Background and related work}\label{sec:background}}

In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.

\hypertarget{sec:fdo}{%
\subsection*{FAIR Digital Object}\label{sec:fdo}}

\hypertarget{sec:next-step-fdo}{%
\subsubsection*{Next steps for FDO}\label{sec:next-step-fdo}}

The FAIR Digital Object Forum (\protect\hyperlink{ref-DMEBhHID}{{`FAIR Digital Objects Forum \textbar{}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public. As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:

\textbf{FAIR Digital Object Overview and Specifications} (\protect\hyperlink{ref-11NpETLq1}{Anders et al., 2022a}) is a comprehensive overview of FAIR Digital Object specifications listed below. It serves as a primer that introduces FDO concepts and the remaining documents.

The \textbf{FDO Forum Document Standards} (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions. In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.

The \textbf{FDO Requirement Specifications} (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) is an update of (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) as the foundational definition of FDO. This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations. The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.

The \textbf{Machine actionability} (\protect\hyperlink{ref-iJeys0T5}{Weiland et al., 2022a}) sets out to define what is meant by \emph{machine actionability} for FDOs. \emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar. The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.

\textbf{Configuration Types} (\protect\hyperlink{ref-SPd0uEqO}{Lannom et al., 2022a}) classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs. Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022})

\textbf{PID Profiles \& Attributes} (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}) specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile. This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}). This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.

\textbf{Kernel Attributes \& Metadata} (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc. This document expands on RDA Recommendation on PID Kernel Information (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}). It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.

\textbf{Granularity, Versioning, Mutability} (\protect\hyperlink{ref-16W9AycSW}{Working Group, 2022b}) considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements. The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.

\textbf{DOIP Endorsement Request} (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) is an endorsement of the DOIP v2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) specification as a potential FDO implementation, as it has been applied by several institutions (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}). The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).

\textbf{Upload of FDO} (\protect\hyperlink{ref-1H7cmmr69}{Blanchi et al., 2022a}) illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve). ResourceSync (\protect\hyperlink{ref-eS5y9TRh}{{`ResourceSync Framework Specification - Table of Contents'}}) is suggested as one type of service to list FDOs. This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).

\textbf{Typing FAIR Digital Objects} (\protect\hyperlink{ref-ZFzPxCNB}{Lannom et al., 2022b}) defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose. This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types. Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work. How to define such FDO operations is not specified.

\textbf{Implementation of Attributes, Types, Profiles and Registries} (\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) details how to establish FDO registries for types and FDO profiles, with their association with PID systems. This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices. Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} (\protect\hyperlink{ref-1CqIZ47pu}{Schwardmann \& Kálmán, 2022}).

It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.

The concept of \textbf{FAIR Digital Objects} (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) has been introduced as way to expose research data as active objects that conform to the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}). This builds on the \emph{Digital Object} (DO) concept (\protect\hyperlink{ref-11MnuwJ4l}{Kahn \& Wilensky, 2006}), first introduced in 1995 (\protect\hyperlink{ref-3Uqe3fuK}{Kahn \& Wilensky, 1995}) as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles. DO was the inspiration for the ITU X.1255 framework (\protect\hyperlink{ref-103Hw8H43}{{`X.1255 : Framework for discovery of identity management information'}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}).

In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}. The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs). The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing (\protect\hyperlink{ref-tz0P3DTC}{Wittenburg et al., 2019}).

Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}), in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}. Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}). The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).

\hypertarget{fdo-approaches}{%
\subsubsection*{FDO approaches}\label{fdo-approaches}}

FDO is an evolving concept. A set of FDO Demonstrators (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) highlight how current adapters are approaching implementations of FDO from different angles:

\begin{itemize}
\tightlist
\item
  Building on the Digital Object concept, using the simplified DOIP v2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), see (\protect\hyperlink{ref-xuQKRRx5}{{`DOIP and Examples --- Cordra documentation'}}).} (usually TCP/IP over TLS). The main DOIP operations are retrieving, creating and updating digital objects. These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}. FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}}) and PIDs are assigned using the Handle system. Several type registries have been established.
\item
  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
\item
  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
\end{itemize}

From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.

\hypertarget{sec:ld}{%
\subsection*{From the Semantic Web to Linked Data}\label{sec:ld}}

In order to describe \emph{Linked Data} as it is used today, we'll start with an (opinionated) description of the evolution of its foundation, the \emph{Semantic Web}.

\hypertarget{sec:semweb}{%
\subsubsection*{A brief history of the Semantic Web}\label{sec:semweb}}

The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee (\protect\hyperlink{ref-Guy7cjVY}{Berners-Lee \& Fischetti, 1999}), at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs. The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}. The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.

Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\protect\hyperlink{ref-16WGmCih8}{{`RDF 1.1 Primer'}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\protect\hyperlink{ref-16wvFH4nh}{{`Resource Description Framework (RDF) Model and Syntax Specification'}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).

While triple-based knowledge representations were not new (\protect\hyperlink{ref-NxvhwSmG}{Stanczyk, 1987}), the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs (\protect\hyperlink{ref-RRtwPF7Y}{Berners-Lee, Fielding \& Masinter, 2005}) are generalised forms of URLs that include locator-less identifiers
  such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years (\protect\hyperlink{ref-nW4HY8Nq}{{`"info" URI Registry (Frozen)'}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) (\protect\hyperlink{ref-HqZ3r6J3}{Juty, Le Novere \& Laibe, 2011}). RDF 1.1 formats use Unicode to support IRIs (\protect\hyperlink{ref-rqjuXkEf}{Duerst \& Suignard, 2005}), which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.

The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data). Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term. The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource. By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs. The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.

The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is). More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} (\protect\hyperlink{ref-yd2WE9eq}{Horrocks \& Hendler, 2002}) gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.

Established Semantic Web services and standards include SPARQL (\protect\hyperlink{ref-xww8fb1B}{{`SPARQL 1.1 Overview'}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).

The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\protect\hyperlink{ref-bMSxa8XW}{Goble \& Stevens, 2008}; \protect\hyperlink{ref-Qlrlr2TZ}{Williams et al., 2012}). In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same. One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.

Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies. Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties). Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) (\protect\hyperlink{ref-FDDmgO6s}{Hu et al., 2011}) .

The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources. This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS (\protect\hyperlink{ref-191ZkYZMt}{Groth et al., 2014})), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps (\protect\hyperlink{ref-9CdJnsoJ}{Kamdar, Tudorache \& Musen, 2017}). Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication (\protect\hyperlink{ref-AGHR9oyr}{Klein et al., 2014}), for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data (\protect\hyperlink{ref-5SuwLNdE}{Polleres et al., 2020, p. handle:20.500.11811/7183}). Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies (\protect\hyperlink{ref-TUmA4KQT}{Carriero et al., 2020}) , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), OBO (\protect\hyperlink{ref-1DalZuKct}{Tirmizi et al., 2011}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})), hash vs slash, HTTP status codes and PID redirection strategies (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).

\hypertarget{sec:ld-web}{%
\subsubsection*{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}

The \textbf{Linked Data} concept (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) was kickstarted as a set of best practices (\protect\hyperlink{ref-OGXYtGZ8}{{`Linked Data - Design Issues'}}) to bring the Web aspect back into focus. Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers. This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.

Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data. In addition to several such embedded \emph{microformats} (Open Graph (\protect\hyperlink{ref-oN9D5LkD}{{`The Open Graph protocol'}}), RDFa (\protect\hyperlink{ref-FKIF6ApB}{{`RDFa 1.1 Primer - Third Edition'}}), Microdata (\protect\hyperlink{ref-14Utscfe4}{{`HTML Standard'}})) we find JSON-LD (\protect\hyperlink{ref-X5SWPqSU}{{`JSON-LD 1.1'}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications. JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}).

Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\protect\hyperlink{ref-BFMoe8FC}{{`Designing a Linked Data developer experience'}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\protect\hyperlink{ref-GDrm9dgX}{{`Shapes Constraint Language (SHACL)'}}) or ShEx (\protect\hyperlink{ref-YUzogaV}{{`Shape Expressions (ShEx) 2.1 Primer'}})) can be used to validate RDF Data (\protect\hyperlink{ref-ZeOAPNHa}{Gayo et al., 2017}; \protect\hyperlink{ref-S5BI6Mwx}{Thornton et al., 2019}) before consuming it programmatically, or reshaping data to fit other models. While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts (\protect\hyperlink{ref-o596f2JM}{Klímek, Škoda \& Nečaský, 2019}), which then tend to prefer non-semantic two-dimensional formats such as CSV files.

A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}). One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies (\protect\hyperlink{ref-MhsCt4P3}{Wolstencroft et al., 2011}), the use of Linked Data as part of REST Web APIs (\protect\hyperlink{ref-eS6a6tlz}{Page, De Roure \& Martinez, 2011}) , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy (\protect\hyperlink{ref-117tX9MEG}{2016}), with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.

\hypertarget{sec:comparing}{%
\section*{Comparing FDO and existing approaches}\label{sec:comparing}}

To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  An Interoperability Framework and Distributed Platform for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}), which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
\item
  The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), validated against its implementations for completeness.
\item
  A Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}), which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
\item
  Cross-checks against RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
\item
  EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
\end{enumerate}

The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (see section \ref{sec:next-step-fdo}).

Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices. For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) against Linked Data in general.

\hypertarget{sec:interoperability-compare}{%
\subsection*{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}

>>>>

\renewcommand*{\arraystretch}{1.4}
